Financial markets for reaching ``green" targets: Cap-and-trade strategies in the Chilean electricity sector

### 2020-04-24 draft_paper/draft_v1.tex
Cambios previos al 24 abril

\begin{abstract}
In the context of the Paris Agreement, Chile has pledged to reduce Greenhouse Gases (GHG) intensity by at least 30\% below 2007 levels by 2030, and to phase out coal as a energy source by 2040, among other strategies. In pursue of these goals, Chile has implemented a \$5 per tonne of CO$_2$ emission tax, first of its kind in Latin America. However, such a low price has proven to be insufficient. In our work, we study an alternative approach for capping and pricing carbon emissions in the Chilean electric sector; the cap and trade paradigm. We model the Chilean electric market (generators and emissions auctioneer) as a two stage capacity expansion equilibrium problem, where we allow future investment and trading of emission permits among generator agents. The model allows to study generation and future investments in the Chilean electric sector in two regimes of demand: deterministic and stochastic. We show that the current Chilean GHG intensity pledge does not drive an important shift in the future Chilean electric matrix. To encourage a shift to greener technologies, a more stringent carbon budget must be considered, resulting in a carbon price approximately ten times higher than the present one. We also show that achieving the emissions reduction goal does not necessarily results in further reductions of carbon generation, or phasing out coal in the longer term. Finally, we demonstrate that under technology change costs reductions, higher demand scenarios will relax the need for stringent carbon budgets to achieve new renewable energy investments and hence meet the Chilean pledges. These results suggest that some aspects of the Chilean pledge require further analysis, of the economic impact, particularly with the recent announcement of achieving carbon neutrality towards 2050. 
\end{abstract}

@@

Our model assumes two basic stages: In the first stage (period $t=0$), electric producers decide their generation, capacity investment and carbon allowance allocation to satisfy an exogenous demand $D(0)\in\mathbb{R}_+$. In the second stage ($t\in T:=\{1,...\bar{t}\}$), uncertainty is revealed in period $t=1$. Uncertainty is represented by a state of nature $\omega\in\Omega:=\{1,..,K\}$. Thus, for a given state of nature $\omega\in\Omega$, the pair $(t,\omega)$ represents the demand in period $t>1$ when the state $\omega$ is reached. For each pair $(t,\omega)$, given an initial allowance allocation in period $t=0$ and a secondary CO$_2e$ permit trading in $t=1$, electric generators engage in a spot market and decide their generation and new capacity investment such that the exogenous stochastic demand level $D(t,\omega)\in\mathbb{R}_+^{T\times\Omega}$ is satisfied. The demand for all periods is denoted by $D=\left(D(0),(D(t,\omega))_{(t,\omega)\in T\times\Omega}\right)\in\mathbb{R}_+\times\mathbb{R}_+^{T\times\Omega}$.

@@

For each generation technology, we obtained the data for capital investment and operation cost (CAPEX and OPEX), installed capacity and future planned capacities expansion through the \textit{Comisi\'on Nacional de Energ\'ia} (CNE) website. Information about the resource potentials were obtained from \cite{Santana2014potencial}.
In the case of the demand, we also use projection provided the CNE\footnote{http://datos.energiaabierta.cl/dataviews/251663/proyeccion-de-demanda-en-gwh/}, which foresees the demand until year 2038. This data is linearly extrapolated to extend the demand projection up to 2050. This demand  is the one used in the deterministic analysis.

@@

The model is calibrated to the above data, considering the operation in 2018. This calibration was performed by fine-tuning a subset of the parameters of the model, such as cost and the technology/investment change, $TC$ and $TCR$ (see Section~\ref{sec:model} for the details). In any case, the estimation of these parameters follows the trends expected for future years (as presented by \cite{Mena2019csp,NRELcosts}), where investment and operation costs of NCRE sources are expected to decrease, as opposed to fuel-based generation where costs tends to remain constant \cite{Mena2019csp,NRELcosts}.
Figure~\ref{fig:bc}(a) shows the electric generation mix produced in our model for 2018. Most of the electricity production of the country comes from coal and gas-fueled generators. Only 18\% of the energy matrix comes from NCRE sources, and taken into account large hydropower plants, this percentage increases up to  46\%. This values are in good agreement with the real operation data from CNE \cite{CNE2018}, which estimates that the contribution from NCREs is 20\% and raises to 50\% when large hydropower plants are included.

@@

Based on Chile's pledge in the Paris Agreement, the projected total emissions, excluding land use, land-use change and  forestry (LULUCF), up until the year 2030 are as shown in Figure~\ref{fig:cap}, based on the most recent data about Chile's NDCs found in the Climate Action Tracker (CAT) (2 Dec 2019) webpage\footnote{https://climateactiontracker.org/countries/chile/}. It should be noted that these emissions consider the main Greenhouse Gases (GHG)-- i.e $CO_2$, $CH_4$ and $N_2O$.  In the Chilean case, the total emissions are dominated by $CO_2$ (78\%), followed by $CH_4$ with 12.5\% and $N_2O$  with 6\% \cite{ChileInformeBienal2016}. The remaining emissions are due to fluoride gases, which we will neglect in the following analysis. To account for all the GHGs, we used emission factors $\epsilon_i$ in units of carbon equivalent tCO$_2e$ taken from \cite{IPCC2014}.

@@

\begin{itemize}
    \item By 2030: 30\% below 2007 GHG intensity of GDP (uncondicional target). This translates in 151\% above 1990 emissions excl. LULUCF by 2030. 
    \item By 2035: 60\% electricity production from renewable energy 
    \item By 2050: 70\% electricity production from renewable energy
    \item By 2024: close eight of the oldest coal-fired power plants (20\% of current coal electricity capacity) 
    \item By 2040: phase-out coal 
\end{itemize}

@@

We now present the effect of an emission cap and the associated emissions trading system in the context of the Chilean electric sector under a deterministic future demand. We perform a sensitivity analysis on the carbon budget estimate presented in Section  \ref{sec:cap}. Note that the mathematical formulation of the Auctioneer's problem allows in its general form to have Normal distributed carbon budget (see $CAP$ in Section \ref{sec:auctioneer}), with mean $\mu$ and variance $\sigma^2$. For simplicity of the analysis, we do not consider a variance ($\sigma=0$) and hence we simply vary the mean $\mu$ parameter between  100 MtCO$_2e$ and 1000 MtCO$_2$e. The pledge carbon budget determined in Section~\ref{sec:cap} falls within this interval (930 MtCO$_2e$).

@@

Figure~\ref{fig:prod_cap} shows production levels per technology towards 2050, for the carbon budget ranging between 100 MtCO$_2e$ and 1000 MtCO$_2e$. The two top panels in  this figure are the closest values to the pledge cap. Generation from conventional fossil-fuel (coal and gas) does not yield in favor of NCREs; moreover, fossil fuels are still important players in the power generation. This result shows that the Chilean pledge of lowering emissions is not sufficient by itself. According to our model, there is little economic incentive (or cost, measured by the price of carbon) for producers to invest in "greener" technologies with a non-restrictive carbon budget. Indeed, the $CO_2e$ price obtained in the less stringent case (1000 MtCO$_2e$) is $\$24/tCO_2e$ versus a price of $\$174/tCO_2e$ in the most stringent case (100 MtCO$_2e$), representing an increment of 6.25 times in the price of carbon permits\footnote{Furthermore, the carbon permit price of $\$24 tCO_2e$ is above the current carbon tax of $\$5 tCO_2$ implemented by the Chilean government.}. 
\smallskip

@@

\subsection{Impact of demand uncertainty on the Chilean pledges}\label{sec:scenarios}
We now consider the case where there is demand uncertainty in future periods. The stochastic demand creates uncertainty with respect to how many allowances generators should get during the first stage, hence, creating different carbon price patterns. We consider five demand scenarios, based on projections from Chile's roadmap towards 2050
\cite{HojadeRuta}\footnote{Note that this is a different entity than the CNE, which provided the real historical demand and its projection used in the case of the deterministic analysis presented in Section \ref{sec:deterministic}}. We also assume that each scenario is equally likely ($Prob_\omega$=1/5). The demand scenarios considered here are \textit{Low, Mid and High GDP, Electrification and High Effort}, as shown in Figure~\ref{fig:demand_scn}. The Electrification scenario assumes a replacement of fuels for electricity in tertiary sectors, such as heating, industries and transportation. The high effort scenario assumes energy efficiency policies in action, resulting in a significant reduction on electricity consumption as compared to the other scenarios.

>>>>

\begin{abstract}
In the context of the Paris Agreement, Chile has pledged to reduce Greenhouse Gases (GHG) intensity by at least 30\% below 2007 levels by 2030, and to phase out coal as a energy source by 2040, among other strategies. In pursue of these goals, Chile has implemented a \$5 per tonne of CO$_2$ emission tax, first of its kind in Latin America. However, such a low price has proven to be insufficient. In our work, we study an alternative approach for capping and pricing carbon emissions in the Chilean electric sector; the cap and trade paradigm. We model the Chilean electric market (generators and emissions auctioneer) as a two stage capacity expansion equilibrium problem, where we allow future investment and trading of emission permits among generator agents. The model allows to study generation and future investments in the Chilean electric sector in two regimes of demand: deterministic and stochastic. We show that the current Chilean Greenhouse Gases (GHG) intensity pledge does not drive an important shift in the future Chilean electric matrix. To encourage a shift to greener technologies, a more stringent carbon budget must be considered, resulting in a carbon price approximately ten times higher than the present one. We also show that achieving the emissions reduction goal does not necessarily results in further reductions of carbon generation, or phasing out coal in the longer term. Finally, we demonstrate that under technology change costs reductions, higher demand scenarios will relax the need for stringent carbon budgets to achieve new renewable energy investments and hence meet the Chilean pledges. These results suggest that some aspects of the Chilean pledge require further analysis, of the economic impact, particularly with the recent announcement of achieving carbon neutrality towards 2050. 
\end{abstract}

@@

Our model assumes two basic stages: In the first stage (period $t=0$), electric producers decide their generation, capacity investment and carbon allowance allocation to satisfy an exogenous demand $D(0)\in\mathbb{R}_+$. In the second stage ($t\in T:=\{1,...\bar{t}\}$), uncertainty is revealed in period $t=1$. Uncertainty is represented by a state of nature $\omega\in\Omega:=\{1,..,K\}$ with probability defined by $Pr(\omega)\in[0,1]$. Thus, for a given state of nature $\omega\in\Omega$, the pair $(t,\omega)$ represents the demand in period $t>1$ when the state $\omega$ is reached. For each pair $(t,\omega)$, given an initial allowance allocation in period $t=0$ and a secondary CO$_2e$ permit trading in $t=1$, electric generators engage in a spot market and decide their generation and new capacity investment such that the exogenous stochastic demand level $D(t,\omega)\in\mathbb{R}_+^{T\times\Omega}$ is satisfied. The demand for all periods is denoted by $D=\left(D(0),(D(t,\omega))_{(t,\omega)\in T\times\Omega}\right)\in\mathbb{R}_+\times\mathbb{R}_+^{T\times\Omega}$.

@@

For each generation technology, we obtained the data for capital investment and operation cost (CAPEX and OPEX), installed capacity and future planned capacities expansion through the \textit{Comisi\'on Nacional de Energ\'ia} (CNE) website. Information about the resource potentials were obtained from \cite{Santana2014potencial}.
In the case of the demand, we also use the projection provided by the CNE\footnote{http://datos.energiaabierta.cl/dataviews/251663/proyeccion-de-demanda-en-gwh/}, which foresees the demand until year 2038. This data is linearly extrapolated to extend the demand projection up to 2050. This demand  is the one used in the deterministic analysis.

@@

The model is calibrated to the above data, considering the operation in 2018. This calibration was performed by fine-tuning a subset of the parameters of the model, such as cost and the technology/investment change, $TC$ and $TCR$ (see Section~\ref{sec:model} for the details). In any case, the estimation of these parameters follows the trends expected for future years (as presented by \cite{Mena2019csp,NRELcosts}), where investment and operation costs of NCRE sources are expected to decrease, as opposed to fuel-based generation where costs tends to remain constant \cite{Mena2019csp,NRELcosts}.
Figure~\ref{fig:bc}(a) shows a contrast of the electric generation mix produced in our model for 2018 with respect to the real operation data. Most of the electricity production of the country comes from coal and gas-fueled generators. Only 18\% of the energy matrix comes from NCRE sources, and taken into account large hydropower plants, this percentage increases up to  46\%. This values are in good agreement with the real operation data from CNE \cite{CNE2018}, which estimates that the contribution from NCREs is 20\% and raises to 50\% when large hydropower plants are included.

@@

Based on Chile's pledge in the Paris Agreement, the projected total emissions, excluding land use, land-use change and  forestry (LULUCF), up until the year 2030 are as shown in Figure~\ref{fig:cap}, based on the most recent data about Chile's NDCs found in the Climate Action Tracker (CAT) (2 Dec 2019) webpage.\footnote{https://climateactiontracker.org/countries/chile/} It should be noted that these emissions consider the main GHG, i.e, $CO_2$, $CH_4$ and $N_2O$.  In the Chilean case, the total emissions are dominated by $CO_2$ (78\%), followed by $CH_4$ with 12.5\% and $N_2O$  with 6\% \cite{ChileInformeBienal2016}. The remaining emissions are due to fluoride gases, which we will neglect in the following analysis. To account for all the GHGs, we used emission factors $\epsilon_i$ in units of carbon equivalent tCO$_2e$ taken from \cite{IPCC2014}.

@@

\begin{itemize}
    \item By 2030: 30\% below 2007 GHG intensity of GDP (uncondicional target). This translates in 151\% above 1990 emissions excl. LULUCF by 2030. 
    \item By 2024: close eight of the oldest coal-fired power plants (20\% of current coal electricity capacity) 
    \item By 2035: 60\% electricity production from renewable energy 
    \item By 2040: phase-out coal 
    \item By 2050: 70\% electricity production from renewable energy
\end{itemize}

@@

We now present the effect of an emission cap and the associated emissions trading system in the context of the Chilean electric sector under a deterministic future demand. We perform a sensitivity analysis on the carbon budget estimate presented in Section  \ref{sec:cap}. Note that the mathematical formulation of the Auctioneer's problem allows in its general form to have Normal distributed carbon budget (see $CAP$ in Section \ref{sec:auctioneer}), with mean $\mu$ and variance $\sigma^2$. For simplicity of the analysis, we do not consider a variance ($\sigma=0$) and hence we simply vary the mean $\mu$ parameter between  100 MtCO$_2e$ and 1000 MtCO$_2$e. The pledge of carbon budget determined in Section~\ref{sec:cap} (930 MtCO$_2e$) belong to that interval.

@@

Figure~\ref{fig:prod_cap} shows production levels per technology towards 2050, for the carbon budget ranging between 100 MtCO$_2e$ and 1000 MtCO$_2e$. The two top panels in  this figure are the closest values to the pledge cap. Generation from conventional fossil-fuel (coal and gas) does not yield in favor of NCREs; moreover, fossil fuels are still important players in the power generation. This result shows that the Chilean pledge of lowering emissions is not sufficient by itself. According to our model, there is little economic incentive (or cost, measured by the price of carbon) for producers to invest in ``greener" technologies with a non-restrictive carbon budget. Indeed, the $CO_2e$ price obtained in the less stringent case (1000 MtCO$_2e$) is $\$24/tCO_2e$ versus a price of $\$174/tCO_2e$ in the most stringent case (100 MtCO$_2e$), representing an increment of 6.25 times in the price of carbon permits.\footnote{Furthermore, the carbon permit price of $\$24 tCO_2e$ is above the current carbon tax of $\$5 tCO_2$ implemented by the Chilean government.} 
\smallskip

@@

\subsection{Targeted Technology Segmentation}

Projections based on the analysis above show that the pledge regarding coal phase out is difficult to honor. Thus, it is necessary to investigate new mechanisms to accomplish the targets.

The cap and trade configuration establish a permits market and no additional trade-restrictions for producers. Unrestricted access to markets is not a realistic setting since market access is ruled by revenue, capacity, caps or quotas among many other regulatory or institutional considerations. For instance, at a general level we analyzed caps that determine the size of the permits market. At an individual level, restricted access can take sophisticated shapes: options, collateral requirements, margin calls, personalized caps and many other financial instruments.

In a competitive setting, (exogenous) restrictions as the one given by a quota were introduced by Cass (1984). Endogenous restrictions are analyzed in Cass et al. (2001), Carosi et al. (2009) among others. Nevertheless, the case of endogenous investment restrictions was not included until CE-TM (2018). Here, we limit a long position in the permits market by a proportion of the permits bought int the first stage. Moreover, the endogenous cap can be targeted to specific technologies (segmentation).

The results of the consideration of such constraints segmenting Coal

\subsection{Impact of demand uncertainty on the Chilean pledges}\label{sec:scenarios}
We now consider the case where there is demand uncertainty in future periods. The stochastic demand creates uncertainty with respect to how many allowances generators should get during the first stage, hence, creating different carbon price patterns. We consider five demand scenarios, based on projections from Chile's roadmap towards 2050
\cite{HojadeRuta}\footnote{Note that this is a different entity than the CNE, which provided the real historical demand and its projection used in the case of the deterministic analysis presented in Section \ref{sec:deterministic}}. We also assume that each scenario is equally likely, $Pr(\omega)=1/5$ for each $\omega\in\Omega$. The demand scenarios considered here are \textit{Low, Mid and High GDP, Electrification and High Effort}, as shown in Figure~\ref{fig:demand_scn}. The Electrification scenario assumes a replacement of fuels for electricity in tertiary sectors, such as heating, industries and transportation. The high effort scenario assumes energy efficiency policies in action, resulting in a significant reduction on electricity consumption as compared to the other scenarios.
